Stochastic Variance Reduced Policy Gradient

### 2018-02-04 svrpg.tex
fixed section labels and svrpg gradient update formula
small fixes + added some references

In a typical SL setting, a performance function $J(\vtheta)$ has to be optimized \wrt to model parameters $\vtheta$. The set of data that are available for training is often a subset of all the data of interest, which may even be infinite, leading to optimization of finite sums to approximate expected performance over an unknown data distribution. When generalization to the complete dataset is not taken into consideration we talk about Empirical Risk Minization (ERM). Even in this case, stochastic optimization is often used for reasons of efficiency. The idea of stochastic gradient descent \cite{nesterov2013introductory} is to iteratively focus on a random subset of the available data to obtain an approximate improvement direction. At the level of the single iteration, this can be much less expensive than taking into account all the data. However, the sub-sampling of data samples is a source of variance that can potentially compromise convergence, so that per-iteration efficiency and convergence rate must be traded off with proper handling of meta-parameters.

@@

Among RL approaches, policy gradient is the one that bears the closest similarity to SL solutions. The fundamental principle of this methods is to optimize a parametric policy by means of stochastic gradient descent. Compared to other applications of SGD, the cost of collecting samples can be very high since it requires to interact with the environment. This makes SVRG-like methods potentially much more efficient than, \eg batch learning.

@@

Unfortunately, RL has a series of difficulties that are not present in ERM. First, in SL the objective can often be designed to be strongly convex. This is not possible in RL, so we have to deal with non-convexities. Then, as mentioned, the dataset is not initially available and may even be infinite, which  makes approximations unavoidable. This also rules out SAG and SAGA because of their storage requirements, which leaves SVRG as the most promising choice. Finally, the distribution used to sample data is not under direct control of the algorithm designer, but is a function of policy parameters that changes over time as the policy is optimized, which is a form of nonstationarity.

% Related work

@@

The main limitation of plain policy gradient has been the high variance of these estimators.
The naive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples by interacting with the environment.
For this reason, literature has focused on the introduction of baselines aiming to reduce the variance~\citep[\eg][]{Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.\footnote{A baseline is a function $f(s,a)$ such that $\EVV[a \sim \pol]{f(s,a)} = 0$, for any $s\in\Sspace$.}
There has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Altough, these techniques has been mainly derived for finite-sum problems, we will show in Sec.~\ref{sec:algo} how can be used in RL.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present SVRG algorithm that is at the core of this work.

\subsection{Stochastic Variance-Reduced Gradient}
Finite-sum optimization is the problem of maximizing an objective function $f(\vtheta)$ which can be decomposed into the sum or average of a finite number of functions $g_i(\cdot|\vtheta)$:
\begin{align*}
        \max_{\vtheta} \left\{ f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g_i(\vtheta)\right\}.
\end{align*}
This kind of optimization is very common in machine learning, where each $g_i$ may correspond to a data sample $x_i$ from a dataset $\mathcal{D}_N$ of size $N$ (\ie $g_i(\vtheta) = g(x_i|\vtheta)$). 
%% In this case, we adopt the following, more meaningful notation:
%% \begin{align*}
%% \max_{\vtheta} f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g(x_i \vert \vtheta).
%% \end{align*}
A common requirement is that $g$ must be smooth and {\color{red} concave} in $\vtheta$.\footnote{Note that we are considering a maximization problem instead of the classical minimization one.} Under this hypothesis, full gradient ascent \cite{cauchy1847methode} with a constant step size achieves an Incremental First-order Oracle (IFO) complexity~\citep{agarwal2014lower} of $O(\nicefrac{N}{\epsilon})$, \ie it requires $k = O(\nicefrac{N}{\epsilon})$ gradient computations to achieve $\norm[]{\nabla_{\vtheta}f(\vtheta_k)}^2\leq\epsilon$.\todopirout{Is it $\ell_2$-norm?} 
This can be too expensive for large values of $N$. Stochastic Gradient (SG) ascent~\citep[\eg][]{robbins1951stochastic,bottou2004large} overcomes this problem by sampling a single sample $x_i$ per iteration, but a vanishing step size is required to control the variance introduced by sampling. The resulting IFO complexity is $O(\nicefrac{1}{\epsilon^2})$ in expectation, which does not depend on $N$, but has a worse dependency on $\epsilon$. 
\todopir{What is the reference?}
Starting from SAG, a series of variations to SG have been proposed to achieve a better trade-off between convergence speed and cost per iteration: \eg SAG~\citep{roux2012stochastic}, SVRG~\cite{johnson2013accelerating}, SAGA~\cite{defazio2014saga} and Finito~\cite{defazio2014finito}, 
The common idea is to reuse past gradient computations to reduce the variance of the current estimate.
In particular, Stochastic Variance-Reduced Gradient (SVRG) is often preferred to other similar methods for its limited storage requirements, which can become a problem when deep and/or wide neural networks are employed.

@@

&+ \frac{1}{B}\left[\sum_{i=0}^{B-1}

@@

\section{Convergence Guarantees}\label{sec:conv}
In this section we adapt existing convergence guarantees for nonconvex SVRG \cite{reddi2016stochastic} \cite{allen2016variance} to the Policy Gradient framework.
Each of the three challenges presented at the beginning of Section \ref{sec:algo} can potentially prevent convergence, so we need additional assumptions:
\begin{enumerate}
\item \textit{Non-convexity}:
	as already mentioned, a common assumption for nonconvex optimization is the L-smoothness of the objective. We actually require the following, which is stronger:
	\begin{assumption}\label{ass:bounded_score}
		For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
		\begin{align*}
		&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
		&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
		\end{align*}
	\end{assumption}
	This property is satisfied by the commonly used truncated Gaussian policy, and is a sufficient condition for the L-smoothness of the objective (see Lemma \ref{lemma:lsmooth} in the appendix).
\item \textit{Approximation}:
since we can only compute an estimate of the full gradient, we require its variance to be bounded:
	\begin{assumption}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for each policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
	\end{assumption}
This assumption is satisfied by the REINFORCE and G(PO)MDP estimators and Gaussian policies \cite{zhao2011analysis} \cite{pirotta2013adaptive}. The problem of full gradient approximation for SVRG has been addressed in the supervised learning framework \cite{harikandeh2015stopwasting} with results similar to ours.
\item \textit{Non-stationarity}:
the variance introduced by importance sampling must be bounded:
	\begin{assumption}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
		\[
		\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS.
		\]
	\end{assumption}
Differently from the previous two, this assumption must be enforced by a proper handling of the epoch size $m$.
\end{enumerate}

@@

\section{Practical Variants}
\todomat{Missing opening to section}
We introduce a term with importance weights to guarantee proper sampling of trajectories.
This term's variance is subjected to the variance of importance weights. 
There are some methods to reduce this variance.

@@

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.
\begin{definition}
For a random variable $X$:
\begin{align*}
	&\mathbb{E}_{t\vert s}\left[X\right] \coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}}$ is the amount of trajectories sampled for the snapshot and ${\it {\bf B}}$ is the amount of trajectories sampled for a sub iteration. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.

>>>>

In a typical SL setting, a performance function $J(\vtheta)$ has to be optimized \wrt to model parameters $\vtheta$. The set of data that are available for training is often a subset of all the cases of interest, which may even be infinite, leading to optimization of finite sums to approximate expected performance over an unknown data distribution. When generalization to the complete dataset is not taken into consideration we talk about Empirical Risk Minization (ERM). Even in this case, stochastic optimization is often used for reasons of efficiency. The idea of stochastic gradient ascent \cite{nesterov2013introductory} is to iteratively focus on a random subset of the available data to obtain an approximate improvement direction. At the level of the single iteration, this can be much less expensive than taking into account all the data. However, the sub-sampling of data samples is a source of variance that can potentially compromise convergence, so that per-iteration efficiency and convergence rate must be traded off with proper handling of meta-parameters.

@@

Among RL approaches, policy gradient \cite{sutton2000policy} is the one that bears the closest similarity to SL solutions. The fundamental principle of this methods is to optimize a parametric policy by means of stochastic gradient ascent. Compared to other applications of SGD, the cost of collecting samples can be very high since it requires to interact with the environment. This makes SVRG-like methods potentially much more efficient than, \eg batch learning.

@@

Unfortunately, RL has a series of difficulties that are not present in ERM. First, in SL the objective can often be designed to be strongly concave. This is not possible in RL, so we have to deal with non-concavities. Then, as mentioned, the dataset is not initially available and may even be infinite, which  makes approximations unavoidable. This also rules out SAG and SAGA because of their storage requirements, which leaves SVRG as the most promising choice. Finally, the distribution used to sample data is not under direct control of the algorithm designer, but is a function of policy parameters that changes over time as the policy is optimized, which is a form of nonstationarity.

% Related work?

@@

The main limitation of plain policy gradient has been the high variance of these estimators.
The naive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples by interacting with the environment.
For this reason, literature has focused on the introduction of baselines aiming to reduce the variance~\citep[\eg][]{Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.\footnote{A baseline is a function $f(s,a)$ such that $\EVV[a \sim \pol]{f(s,a)} = 0$, for any $s\in\Sspace$.}
There has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Altough these techniques has been mainly derived for finite-sum problems, we will show in Sec.~\ref{sec:alg} how they can be used in RL.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

\subsection{Stochastic Variance-Reduced Gradient}
Finite-sum optimization is the problem of maximizing an objective function $f(\vtheta)$ which can be decomposed into the sum or average of a finite number of functions $g_i(\cdot|\vtheta)$:
\begin{align*}
        \max_{\vtheta} \left\{ f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g_i(\vtheta)\right\}.
\end{align*}
This kind of optimization is very common in machine learning, where each $g_i$ may correspond to a data sample $x_i$ from a dataset $\mathcal{D}_N$ of size $N$ (\ie $g_i(\vtheta) = g(x_i|\vtheta)$). 
%% In this case, we adopt the following, more meaningful notation:
%% \begin{align*}
%% \max_{\vtheta} f(\vtheta) = \frac{1}{N}\sum_{i=1}^{N}g(x_i \vert \vtheta).
%% \end{align*}
A common requirement is that $g$ must be smooth and {\color{red} concave} in $\vtheta$.\footnote{Note that we are considering a maximization problem instead of the classical minimization one.} Under this hypothesis, full gradient ascent \cite{cauchy1847methode} with a constant step size achieves an Incremental First-order Oracle (IFO) complexity~\citep{agarwal2014lower} of $O(\nicefrac{N}{\epsilon})$, \ie it requires $k = O(\nicefrac{N}{\epsilon})$ gradient computations to achieve $\norm[]{\nabla_{\vtheta}f(\vtheta_k)}^2\leq\epsilon$.\todopirout{Is it $\ell_2$-norm?} 
This can be too expensive for large values of $N$. Stochastic Gradient (SG) ascent~\citep[\eg][]{robbins1951stochastic,bottou2004large} overcomes this problem by sampling a single sample $x_i$ per iteration, but a vanishing step size is required to control the variance introduced by sampling. The resulting IFO complexity is $O(\nicefrac{1}{\epsilon^2})$ in expectation, which does not depend on $N$, but has a worse dependency on $\epsilon$ \cite{nemirovskii1983problem}\cite{reddi2016stochastic}. 
Starting from SAG, a series of variations to SG have been proposed to achieve a better trade-off between convergence speed and cost per iteration: \eg SAG~\citep{roux2012stochastic}, SVRG~\cite{johnson2013accelerating}, SAGA~\cite{defazio2014saga} and Finito~\cite{defazio2014finito}, 
The common idea is to reuse past gradient computations to reduce the variance of the current estimate.
In particular, Stochastic Variance-Reduced Gradient (SVRG) is often preferred to other similar methods for its limited storage requirements, which can become a problem when deep and/or wide neural networks are employed.

@@

&\blacktriangledown J(\vtheta) = \frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \vert \tilde{\vtheta})\Reward(\tau_j) +\\
&\quad\frac{1}{B}\left[\sum_{i=0}^{B-1}

@@

\section{Convergence Guarantees}\label{sec:conv}
In this section we adapt existing convergence guarantees for nonconvex SVRG \cite{reddi2016stochastic} \cite{allen2016variance} to the Policy Gradient framework.
Each of the three challenges presented at the beginning of Section \ref{sec:alg} can potentially prevent convergence, so we need additional assumptions:
\begin{enumerate}
\item \textit{Non-convexity}:
	as already mentioned, a common assumption for nonconvex optimization is the L-smoothness of the objective. We actually require the following, which is stronger:
	\begin{assumption}\label{ass:bounded_score}
		For each trajectory $\tau$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
		\begin{align*}
		&\left|\nabla_{\theta_i}\log\pi_{\vtheta}(\tau)\right| \leq \GRADLOG \\
		&\left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(\tau)\right| \leq \HESSLOG
		\end{align*}
	\end{assumption}
	This property is satisfied by the commonly used truncated Gaussian policy, and is a sufficient condition for the L-smoothness of the objective (see Lemma \ref{lemma:lsmooth} in the appendix).
\item \textit{Approximation}:
since we can only compute an estimate of the full gradient, we require its variance to be bounded:
	\begin{assumption}\label{ass:REINFORCE}
		There is a constant $V<\infty$ such that, for each policy $\pol$:
		\[
		\Tr\mathbb{V}ar\left(\gradApp{\vtheta}{1}\right) \leq \VARRF.
		\]
	\end{assumption}
This assumption is satisfied by the REINFORCE and G(PO)MDP estimators and Gaussian policies \cite{zhao2011analysis} \cite{pirotta2013adaptive}. The problem of full gradient approximation for SVRG has been addressed in the supervised learning framework \cite{harikandeh2015stopwasting} with results similar to ours.
\item \textit{Non-stationarity}:
the variance introduced by importance sampling must be bounded:
	\begin{assumption}\label{ass:M2}
		There is a constant $W<\infty$ such that, for each pair of policies encountered in Algorithm \ref{alg:svrpg} and for each possible trajectory,
		\[
		\mathbb{V}ar\left(\omega(\tau)\right) \leq \VARIS.
		\]
	\end{assumption}
Differently from the previous two, this assumption must be enforced by a proper handling of the epoch size $m$.
\end{enumerate}

@@

\section{Practical Variants}\label{sec:prac}
\todomat{Missing opening to section}
We introduce a term with importance weights to guarantee proper sampling of trajectories.
This term's variance is subjected to the variance of importance weights. 
There are some methods to reduce this variance.

@@

\subsection*{Definitions}
We give some additional definitions which will be useful in the proofs.
\begin{definition}
For a random variable $X$:
\begin{align*}
	&\mathbb{E}_{t\vert s}\left[X\right] \coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}} = [0,\dots,N)$, and ${\it {\bf B}} = [0,\dots,B)$. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.
